A generalization of the Bradley-Terry model for draws in chess

### 2020-09-30 inst/jebo_chess_rebuttal.txt

This interpretation of Moul and Nye is very close to my own, and
re-reading my work I realise that the sentence in question was poorly
phrased and indeed unintentionally misleading.  What I wanted to say
was that chess draws have been studied by many workers and Moul and
Nye observe that

I have rephrased the sentence in question, and added a few more points
of reference to Mouland Nye, in BLUE.

@@

> 4) The paper concludes that the estimates derived from the 1962
> Interzonals and the 1963 Candidates strongly suggest that the "draw
> monster" won games involving two Soviet players more frequently than
> games that involved at least one non-Soviet.  This is consistent with
> a Soviet cartel, but it could also be the result of Soviet-specific
> styles of play.  Moul and Nye (2009) test the Soviet style-of-play
> hypothesis by considering international tournaments when the format
> changed from all-play-all to knockout and all Soviet tournaments.
> Both comparisons yield results consistent with a Soviet draw-cartel in
> international all-play-all tournaments and rejecting the Soviet
> style-of-play hypothesis.  By only considering tournaments that were
> international and all-play-all, the author forgoes this critical
> source of identification.
> 
> In the cartel example, the author also neglects the advantage that
> presumably accrued to Soviet players when they played non-Soviet
> players in all-play-all international tournaments.  This would suggest
> another "monster" to be estimated, one that benefits the Soviet player
> when playing a non-Soviet opponent.  Without such an addition, Soviet
> strengths will be overstated.
>

(comments addressing this comment in DARK GREEN)
This is an excellent suggestion, one which I feel nicely rounds out
the analysis.  I have included a "rest monster" whose strength aids
players following a collusive draw.  Reified Bradley-Terry allows one
not only to establish the existence of such a rest monster, but also
to quantify its strength and I have done just this in the revision.

### 2020-10-01 inst/jebo_chess_rebuttal.txt

Also, I have discussed Formula 1 motor racing above under part 1 of
the review.

> 3) As the author notes, the work of Moul and Nye (2009) is directly on
> point.  The author's discussion of this paper, however, is overly
> sparse and at times misleading.  Moul and Nye use existing observed
> player strengths (variations of commonly used scores based on the Elo
> rating system that goes back to 1960).  Moul and Nye then use these
> strengths within an ordered probit to estimate parameters that relate
> back to draw likelihood and White's first-mover advantage.  The sharp
> difference between that paper and this is then that the author is
> estimating player relative strengths (conditional upon a pool of
> players) while Moul and Nye use "observed" strengths [footnote: Those
> relative strengths are highly predictive, with t-statistics of around
> 20].  The latter approach is far less parametrically demanding, but
> leans much more on the assumption of normally distributed disturbances
> (basically the differing assumption of functional form).  The latter
> approach also allows individuals' player strength to vary across time
> and is far more flexible to changes in the composition of the pool of
> players.
> 
> Furthermore, Moul and Nye estimate several versions of their ordered
> probit without data on the number of moves (ply-count).  This is
> contrary to the implication of the paper (p. 6, immediately below
> Table 2), which suggests that the ply-count data are essential to that
> paper's findings.  All of that paper's core results (significance of
> draws, first-mover advantage, a distinct draw outcome for round-robin
> international tournament games that involve two Soviets) are
> originally derived without ply-count data.  The ply-count data are
> incorporated to explore a potential mechanism by which players in a
> draw-cartel might benefit.
>

This interpretation of Moul and Nye is very close to my own, and
re-reading my work I realise that the sentence in question was poorly
phrased and indeed clearly (but unintentionally) misleading.  I have
rephrased the sentence in question, and added a few more points of
reference to Moul and Nye, in BLUE.

### 2020-10-01 inst/jebo_chess_revision2.tex
more accounting for comments on Moul and Nye (blue)

The seminal statistical analysis of 
\citet{moul2009} consider chess draws and use
sophisticated regression techniques to analyse games in terms of not
only the result, but also the ply-count, data not used in the present
analysis.

@@

\citet{moul2009} consider this accusation from a statistical
perspective and present a detailed analysis based on number of moves
in each game, reasoning that hard-fought games which happened to end
in a draw are more likely to last longer than collusive or
pre-arranged draws.  Their analysis incorporated a dataset of 7355
games played in the period 1940-1978, used to calibrate the ability of
the players.  In contrast, the analysis presented here uses only game
outcomes (win-loss-draw) in the competition itself.

@@

\cite{moul2009} present detailed Monte Carlo simulations to assess the
likely impact of a putative Soviet cartel on tournament outcomes.  In
principle, the present system of likelihood and reified entities would
admit such simulations and it might be interesting to compare results.

\subsection*{Supplementary material}

\subsection*{\textcolor{DarkGreen}{The ``rest monster'' is due to a suggestion by an insightful JEBO referee}.}

>>>>

\textcolor{blue}{The seminal work of \citet{moul2009} presents
  sophisticated regression techniques to analyse draws and first-mover
  advantage.  They compare all-play-all events with knockout format
  events to test---and reject---hypotheses of a distinct Soviet
  style-of-play; the work I present also considers Soviet proclivity
  to draw, using data from a single event, specifically the Stockholm
  Interzonal tournament of (1962).}

@@

\textcolor{blue}{\citet{moul2009} consider this accusation from a
  statistical perspective.  Their analysis incorporated a dataset of
  7355 games played in the period 1940-1978, used to calibrate the
  ability of the players.  In contrast, the analysis presented here
  uses only game outcomes (win-loss-draw) in the competition itself.}

@@

\cite{moul2009} present detailed Monte Carlo simulations to assess the
likely impact of a putative Soviet cartel on tournament outcomes.  In
principle, the present system of likelihood and reified entities would
admit such simulations and it might be interesting to compare results.
\textcolor{blue}{\cite{moul2009} go on to compare round robin
  tournaments to knockout format events, searching for evidence of a
  Soviet draw cartel.  By only considering international tournaments
  [where Soviets have incentive to play differently against fellow
    Soviets or Westerners] the analysis presented here does not allow
  me to do this but the technique could easily be applied to a more
  informative dataset.}

\textcolor{DarkGreen}{The Bradley-Terry techniques used here furnish
  evidence that collusive games played during during Cura\c{c}ao 1962
  were less detrimental to future performance than noncollusive games.
  However, there was no cognitively realistic model for mental fatigue
  (for example, downweighting heavy workloads long in the past) and
  the criterion for restedness itself was quite coarse. not allowing
  It would be possible in principle to investigate the effect of
  recovery using similar reified entities.}

\subsection*{\textcolor{DarkGreen}{Acknowledgement}}

\textcolor{DarkGreen}{The ``rest monster'' concept is due to a
  suggestion by an insightful JEBO referee}

\subsection*{Computational resources, supplementary material}
